# Building interactive plots using `plotly` package!
#### Eva Pelaez
#### 26 Nov 24

## Tutorial Aims

#### <a href="#section1"> 1. Good Data Visualisation</a>

#### <a href="#section2"> 2. Are interactive plots that much better?</a>

#### <a href="#section3"> 3. Introduction to `Plotly` package</a>

#### <a href="#section4"> 4. Scatter plots</a>

#### <a href="#section5"> 5. Bar Charts</a>

#### <a href="#section6"> 6. Choropleths</a>



<a name="section1"></a>

## 1. Good Data Visualisation
We have already learned from our previous [data visualisation tutorials](https://ourcodingclub.github.io/tutorials/datavis/) what makes a good plot. A quick recap on the main points:

1. Appropriate plot type for results

2. Plot is well organised

3. X and Y axis are easy to read and have the correct units

4. Clear and informative legend

<a name="section2"></a>

## 2. Are interactive plots that much better?

You may be asking yourself, why I am putting myself through more coding when I already know how to make plots using the `ggplot2` package? That is because interactive plots are amazing in visualising data in a way that is really engaging.

Interactive plots invite people to actively explore the data being presented to them rather than just looking at a static image. This helps with retention of what the plot is explaining as non-expert audiences can filter and custom the image and build their understanding.

The one downside of interactive plots its that they are not appropriate for academic papers but they are commonly used for online websites and newspapers. So let's get started and learn how to make them!

<a name="section3"></a>

## 3. Introduction to `Plotly` package

This tutorial will consist on learning how to make interactive plots using the `Plotly` package. The `Plotly` package utilities the power of Plotly.js, a popular JavaScript library for interactive plotting, and allow us to use R to create customisable, interactive, web-based visualisations.

Some key features of this package:

 - Interactive plots: zoom, pan and hover features, with clickable legends and data point selection.
 
 - Wide range of plot types: can create any of the basic plots (scatter, bar charts, histograms), as well as more advanced and specialised plots (choropleths, 3D scatter plots, financial charts).
 
 - It works well with other common R packages such as `ggplot2`, `dplyr` or `shiny`.
 
 - It outputs plots that are HTML and web compatible
 
 - Cross-language compatibility: Though it uses R, the underlying Plotly.js library ensures that visualizations are consistent across Python, JavaScript, and R.

This is a short summary of what this package can do but if you are curious on the extent of what this package can do, head over to their [website](https://plotly.com/r/).

During this tutorial will be using `Plotly` to create several scatter plots, bar charts and choropleths and at the same time learning the basic functions of this amazing package. 

<a name="section4"></a>

## 4. Scatter plots
As everyone already knows, before we even start we need to load our libraries!

```{r message=FALSE, warning=FALSE}
#Loading libraries
library(tidyverse)
library(dplyr)
library(ggplot2)

#install.packages("plotly")
library(plotly)

```

We will use data from the Living Planet Index, which you have already downloaded from [the Github repository](https://github.com/ourcodingclub/CC-4-Datavis) (Click on Clone or Download/Download ZIP and then unzip the files)

```{r}
# Import data from the Living Planet Index - population trends of vertebrate species from 1970 to 2014
load("data/LPI_data.Rdata")
```

For the first plot, we will be creating a scatter plot using `ggplot2`as a way to then later compare it to a scatter plot created using `plotly`.
We will be plotting the __Elephant abundance from the year 1970 to 2013 in different African regions__

So first we need to reshape and clean up our data:

>**Note:** This tutorial is not focusing on cleaning and manipulating data using `tidyverse`or `dlpyr`. However, if you are still strugling with this, then head over to this [tutorial](https://ourcodingclub.github.io/tutorials/data-manip-intro/) and try it out!

```{r}
# Reshaping data to long format and converting columns to numeric
data2 <- data %>%
  gather("year", "abundance", 25:69) %>%
  mutate(
    year = parse_number(year),
    abundance = as.numeric(abundance)
  )
  
#Cleaning up the data
elephant_clean <- data2 %>%
  filter(str_detect(Common.Name, "African elephant")) %>%
  select(Country.list, year, abundance) %>%
  filter(!Country.list %in% c("Botswana, Tanzania, United Republic Of, South Africa, Uganda, Zimbabwe")) %>%
  na.omit() %>% 
  mutate(Country.list = case_when(
    Country.list == "Tanzania, United Republic Of" ~ "Tanzania",
    Country.list == "Congo, The Democratic Republic Of The" ~ "Dem. Rep. Congo",
    Country.list == "Central African Republic" ~ "Central African Rep.",
    Country.list == "Swaziland" ~ "eSwatini",
    TRUE ~ `Country.list`
  ))
```

To be able to plot the elephant abundance across different African regions we are missing a column on regions. That is why we need to download the following package and data frame to be able to match the countries in our Living Planted Index data set to their respective African regions.

```{r}
#install.packages("rnaturalearth")
#install.packages("rnaturalearthdata")
library(rnaturalearth)                  #These two packages will help us download data for countries and regions

# Load country data from rnaturalearth
countries <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

# Subset to African countries
africa_countries <- countries %>%
  filter(region_un == "Africa") %>%
  select(name, region_un, subregion) %>%
  rename(Country = name, Region = subregion)

#Join our two datasets
elephant_clean <- elephant_clean %>%
  left_join(africa_countries, by = c("Country.list" = "Country"))
  
#Remove the  Western Africa region as there is only 4 data points on it and rename our other regions to make future coding easier 
elephant_clean <- elephant_clean %>% 
  filter(!Region %in% c("Western Africa")) %>% 
  mutate(Region = case_when(
    Region == "Eastern Africa" ~ "Eastern_Africa",
    Region == "Middle Africa" ~ "Middle_Africa",
    Region == "Southern Africa" ~ "Southern_Africa",
    TRUE ~ `Region`
  ))

```

 
## Static scatter plot using `ggplot2`

We already know how to make scatter plots using `ggplot2`, so let's build one really quick.

>**Note:** if this sounds and looks overwhelming (trust me it can be) then head over to the Coding club tutorial on [data visualisation](https://ourcodingclub.github.io/tutorials/datavis/)!

```{r}
(ggplot_scatter <- ggplot(elephant_clean, aes(x= year, y= abundance, colour = Region)) +
    geom_point(size = 3) +
    geom_line() +
    theme_classic() +
    labs(
      title = "Elephant abundance from 1970 to 2013",
      x = "Year",
      y = "Abundance"
    ))
```

Pretty simple to build a non-interactive plot but it is kind of boring now that we know interactive plots exist!

```{r}
#Saving the plot
ggsave("ggplot_scatter.png", ggplot_scatter,width=14,height=8,dpi=450)
```

Now that we have this, lets build an interactive scatter plot!

## Interactive scatter plot

Let's use the same data set to build a basic scatter plot with `plotly`!

```{r}
plot1 <- plot_ly(data = elephant_clean, x= ~year, y= ~abundance, type = "scatter", mode = "markers")
```

The `plot_ly()` function is the main function inside the `plotly` package to build plots. We then have to specify which data set we are using and which variables we are putting in the x and y axis. `type = "scatter"` will tell our function we are wanting to plot a scatter graphs and `mode = "markers"` will tell it that we want dots for our data points.

>**Note:** Whenever you are calling any columns from a data frame using the `plot_ly()` function you need to put the `~` beforehand. 

```{r}
plot1
```

And that is a simple and easy way to build an interactive plot. 

Since the plot is interactive it cannot be saved the same way as static plots (jnep or png will not keep the interactive aspect of the plot), so we need to save it as a .html file! 

```{r}
#Saving interactive plot
htmlwidgets::saveWidget(plot1, "int_scatter1.html")
```

#### Grouped scatter plot
The scatter plot built with `ggplot2` still looks slightly better than the interactive one and its because the population of elephants is grouped by african region, so lets do that really quickly!

```{r}
plot2 <- plot_ly(data = elephant_clean, x= ~year, y= ~abundance, color = ~Region, type= "scatter", mode = "markers")
```

Pretty self explanatory, `color = ~Region`, colours the markers depending on the region. 

>**Note:** sorry to anyone that is used to British spelling but `plotly` will only accept the American spelling; `color` instead of `colour`. 

```{r}
plot2
```

```{r}
htmlwidgets::saveWidget(plot2, "int_scatter2.html")
```

#### Customisation of scatter plot
You may be asking yourself: What if I want to add lines, change the colour of the graph or add labels and titles? Well, lets run throught all of these.

```{r}
plot3 <- plot_ly(data = elephant_clean, x= ~year, y= ~abundance, color = ~Region, type= "scatter", mode = "lines+markers",
                 colors = "Paired") %>% 
  layout(title = 'Elephant Abundance over the years in different regions of Africa',
         yaxis = list(title = "Elephant Abundance"),
         xaxis = list(title = "Year"))
```

Lets break this down, to add lines to the plot you just need to update `mode =` so it tells the function we want both lines and markers. To change the colour of the dots we add the `colors =`function and choose a colorway. Finally to add titles and labels we need to pipe the new function `layout()`. `title = ""` lets you add a title to the plot, `yaxis = list(title = "")` lets you rename the y axis and `xaxis = list(title = "")` lets you rename the x axis.

>**Note:** Be careful with `color =` and `colors =`, even though they look the same, the former lets you group and colour based on a category from your data set but the second one let's you actually change the color of the markers. I know, pretty annoying!

>**Note:** If you want to know all about colors in `plotly`then head over to this [page on their website](https://plotly.com/r/discrete-color/)

```{r}
plot3

```

```{r}
htmlwidgets::saveWidget(plot3, "int_scatter3.html")
```

And with that you have built your first interactive scatter plot using the `plotly` package! Not so scary anymore. If you are curious and want to learn more about building all types of scatter plots with `plotly`then head over to their [website](https://plotly.com/r/line-and-scatter/)!

As we can see a scatter plot may have not been the best choice to represent this data! So lets try build a bar chart instead!


<a name="section5"></a>

## 5. Bar charts
Let's say we want to plot __Elephant abundance over two time ranges in different African regions__. Well a bar chart would be a great way of visualising this!

To do this we need to calculate the mean abundace over the time ranges of our choosing.

```{r}
# Calculate mean abundance for each region in the specified year ranges
mean_abundance <- elephant_clean %>%
  mutate(YearRange = case_when(
    year >= 1970 & year <= 1989 ~ "1970-1989",
    year >= 1990 & year <= 2010 ~ "1990-2010",
    TRUE ~ NA_character_           
  )) %>%
  filter(!is.na(YearRange)) %>%     
  group_by(Region, YearRange) %>%  
  summarize(MeanAbundance = mean(abundance, na.rm = TRUE), .groups = 'drop')

mean_abundance$YearRange <- factor(mean_abundance$YearRange, levels = c("1970-1989", "1990-2010"))
```

## Static bar chart using `ggplot2`
Let's first plot a static bar chart using `ggplot2` so we can then compare it to an interactive one.

```{r}
(ggplot_bar <- ggplot(mean_abundance, aes(fill=Region, x= YearRange, y= MeanAbundance)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Mean Elephant Abundance by Region and Year Range",
       x = "Year Range",
       y = "Mean Abundance",
       fill = "Region") +
  theme_bw())
```

Pretty simple bar chart; but lets be honest, quite boring now that we know how to build interactive plots!

```{r}
ggsave("ggplot_bar.png", ggplot_bar,width=14,height=8,dpi=450)

```

## Interactive bar chart
So there is actually two ways we can build a bar chart with `plotly`. The first one will create our typical bar chart; while the second one will allow us to customize it as we see fit. Bear with me as we go through both ways!

### 1st way!
For the first way, we will build the bar chart the same way as our scatter plot.

```{r}
bar_plotly <- plot_ly(data = mean_abundance,x= ~YearRange, y= ~MeanAbundance, type="bar",color = ~Region, colors = "Paired")
```

We can see the same format as before; `plot_ly` to build the figure, calling the dataset as well as the x and y variables and finally we are still grouping by `color = ~Region` and choosing a specific colorway `colors = ""`. The only difference between this plot and the scatter graph is, that we are telling our function that we want to build a bar chart, `type = "bar"`.

```{r}
bar_plotly
```

```{r}
htmlwidgets::saveWidget(bar_plotly, "simpleway_bar_plotly.html")
```

And there we have it! Easy and simple way to build a bar chart using `plotly`. However, this way doesn't allow us to individually customize our bars or even change the layout of them. This is where the second way comes in handy!

### 2nd way!
First, before we start building our plot, we need to change our data set from the long format to a wide format (counter-intuitive I know, but there is a reason)

```{r}
mean_abund_wide <- spread(mean_abundance,Region, MeanAbundance)
```

This way you should have a column with YearRange and 3 other columns for the different regions and their respective abundance. Now we are ready to create our bar chart!

```{r} 
bar_2way <- plot_ly(data = mean_abund_wide, x = ~YearRange, y = ~Eastern_Africa, type = 'bar', name = 'Eastern Africa') %>% 
  add_trace(y = ~Middle_Africa, name = 'Middle Africa') %>% 
  add_trace(y = ~Southern_Africa, name = 'Southern Africa') %>% 
  layout(yaxis = list(title = 'Abundance'), barmode = 'group')
```

Don't panic! Lets go through the code!
In the first line, we use `plot_ly` to tell R that we are building a plot, using the data from `mean_abund_wide`. We then tell it we want `YearRange`as our x-axis and `Eastern_Africa` as our y-axis (in our first layer) and that we are building a bar chart (`type= 'bar'`). The `add_trace()` function is adding two more layers bars for each of the Regions, each with their respective y value and their respective `name =""`. We are finally telling R that we want our y-axis label to be 'Abundance' and for our bar chart to be grouped `barmode = "group"`.

```{r}
bar_2way
```

```{r}
htmlwidgets::saveWidget(bar_2way, "bar_2nd_way.html")
```

So you may be asking yourself, why go to the trouble of creating a bar chart using the second way if we get the exact same graph? 

This is because if we want to customise the figure its so much easier to do with the second type of code.

#### Stacked bar chart
Say we want to create a stacked bar chart instead.

```{r}
bar_stack <- plot_ly(data = mean_abund_wide, x = ~YearRange, y = ~Eastern_Africa, type = 'bar', name = 'Eastern Africa') %>% 
  add_trace(y = ~Middle_Africa, name = 'Middle Africa') %>% 
  add_trace(y = ~Southern_Africa, name = 'Southern Africa') %>% 
  layout(yaxis = list(title = 'Abundance'), barmode = 'stack')
```

It is really simple, we just need to change our `barmode = ""` from `group` to `stack`.

```{r}
bar_stack
```

```{r}
htmlwidgets::saveWidget(bar_2way, "bar_stack.html")
```

#### Direct labels on bars and rotated x-axis labels
Say we want to include direct labels on each individual bar so its easy to read its value AND that we want our x-axis labels to be rotated, well its really straight forward.

```{r}
direct_rotated_bar <- plot_ly(data = mean_abund_wide, x = ~YearRange, y = ~Eastern_Africa, type = 'bar', name = 'Eastern Africa', 
               text = ~Eastern_Africa, textposition = "auto") %>% 
  add_trace(y = ~Middle_Africa, name = 'Middle Africa', text = ~Middle_Africa, textposition = "auto") %>% 
  add_trace(y = ~Southern_Africa, name = 'Southern Africa', text = ~Southern_Africa, textposition = "auto") %>% 
  layout(xaxis = list(title = "Year Range", tickangle = -45),
                        yaxis = list(title = "Abundance"),
                        margin = list(b = 100),
                        barmode = 'group')
```

To create the direct labels for the individual bars we just need to add `textposition = "auto"` in each line of code that creates a bar and to rotate the x-axis we just add `tickangle =-45` into the line of code that customises the x-axis.

```{r}
direct_rotated_bar
```

```{r}
htmlwidgets::saveWidget(direct_rotated_bar, "direct_rotated_bar.html")
```

#### Specific coloured bars
Finally, say that we wish to choose specific colours for each of the bars.

```{r}
final_bar <- plot_ly(data = mean_abund_wide, x = ~YearRange, y = ~Eastern_Africa, type = 'bar', name = 'Eastern Africa',
                marker = list(color = "rgb(255, 127, 80)")) %>% 
  add_trace(y = ~Middle_Africa, name = 'Middle Africa',
            marker = list(color = 'rgb(135, 206, 235)')) %>% 
  add_trace(y = ~Southern_Africa, name = 'Southern Africa',
            marker = list(color = "rgb(152, 251, 152)")) %>% 
  layout(xaxis = list(title = "Year Range"),
         yaxis = list(title = "Abundance"),
         margin = list(b = 100),
         barmode = 'group')
```

We just need to include `marker = list(color = "")` with specific rgb colour code in each line of code that creates a bar. 

```{r}
final_bar
```

```{r}
htmlwidgets::saveWidget(final_bar, "final_bar.html")
```

And there we have it! We are now able to create interactive bar chart using the `plotly`package and they sure look pretty!

We are almost finished but before we go, lets learn how to create one final type of plot.

<a name="section6"></a>

## 6. Choropleths
Just in case you don't know a choropleth is, they are a type of map used to represent data through varying shades or patterns within predefined geographic regions. 

Let us now create a interactive choropleth of Africa to show us the __Elephant abundance for different countries over the years__.

So lets get coding!

We first need to clean up the data a bit more because some countries have several values of abundance for the same year.

```{r}
elephant_map_df <- elephant_clean %>% 
  group_by(year, Country.list) %>%
  summarize(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop")
```

Lets build the choropleth!

```{r}
africa_map <- plot_geo(elephant_map_df, locationmode = "country names", frame = ~year) %>% 
  add_trace(locations = ~Country.list, 
            z =~total_abundance)
```

A quick explanation of the code. We can see that this time we are using the `plot_geo()` function to build the figure because this function is specifically designed to create geographic visualisations. `locationmode = "country names"` tells R that we are going to be using country names to build our geographic data. The slider at the bottom of the interactive plot that goes through the years is created by `frame = "~year"`. `add_trace()` adds another layer to our plot with `locations = ~Country.list` which maps the countries from our data set and `z=~total_abundance` uses the values in that column to colour or map. 

```{r}
africa_map
```

Its a really fun plot already but lets fix it up. We want to have the focus of the map be in Africa, as all our data is from there, and we want to make sure that the colour scale doesn't change over the years (it can make reading a plot really confusing). Additionlly, lets choose a pretty colour scale!

#### Cutomisation of a choropleth

```{r}
africa_map <- plot_geo(elephant_map_df, locationmode = "country names", frame = ~year) %>% 
  add_trace(locations = ~Country.list, 
            z =~total_abundance,
            zmin = 0,
            zmax = max(elephant_map_df$total_abundance),
            color = ~total_abundance,
            colorscale = "Plasma") %>% 
  layout(geo = list(scope = "africa"),
         title = "Total elephant abundance in Africa\n from 1970 to 2013")
```

Lets run through this code together. `zmin` and `zmax` sets the minimum and maximum values for the color scale: from 0 to the maximum value of total_abundance. `color = ~total_abundance` specifies that the total_abundance column should determine the colors. `colorscale = "Plasma"` applies the predefined Plotly color scale "Plasma". Finally, `geo = list(scope = "africa")` limits the map's geographic scope to Africa and we added a `title =` to the map.

```{r}
africa_map
```

```{r}
htmlwidgets::saveWidget(africa_map, "interactive_map.html")
```

And there we go! Last plot completed!

<a name="section7"></a>

## 7. Challenge yourself!
If you still more energy to keep on coding, play around with some other functions and build other types of interactive plots using the `plotly`package!


## Tutorial outcomes
After this tutorial you have hopefully achieved these learning outcomes:

1. Learnt some basic functions from the `plotly` package

2. Learnt how to plot interactive plots (scatter plots, bar charts, and choropleth plots)

3. Learnt how to change the aesthetics of the plots

## Other resources

Here is the [website for `plotly`](https://plotly.com/r/) with everything you would need to know about this package!










